Numerical Evolution from the Hamiltonian

### 2020-09-24 paper/paper.tex

\documentclass{iopart}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{bbm}
\expandafter\let\csname equation*\endcsname\relax
\expandafter\let\csname endequation*\endcsname\relax
\usepackage{amsmath}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\undertilde}[1]{\underset{\tilde{}}{#1}}
\newcommand{\hmatrix}{\ensuremath{%
  \begin{pmatrix}
    \mathbbm{O}&\mathbbm{I}\\-\mathbbm{I}&\mathbbm{O}
  \end{pmatrix}
}}

@@

\submitto{\EJP}
%\maketitle
\section{Introduction}
In the hamiltonian formulation of classical mechanics, the time evolutin of
physical systems is governed by Hamilton's equations which, for point particle
systems, take the form
\begin{align}
  \dot q_i&=\pd{H}{p_i}&
  \dot p_i&=-\pd{H}{q_i},
\end{align}
for $i=1,\;\ldots,\;F$ where $F$ is the number of degrees of freedom of the
system (three times the number of particles for unconstrained, three dimensional
systems), $q_i$, $p_i$ are the canonical coordinates and conjugate momenta
respectively, and $H=H(t, q_i, p_i)$ is the hamiltonian function of the system
(for XXX systems, the hamiltonian function is simply the mechanical energy of
the system). For discrete systems, these equations are a set of first order
ordinary differential equations (ODE) in time, whose solutions for given initial
conditions are the trajectories folowed by the system's particles.

@@

%-------------------------------------------------------------------------------
\section{Numerical procedure}
%The standard, reasonable approach to the problem of finding the trajectories of
%$N$ interacting point particles using Hamilton's equations is as follows: one
%first derives analytical expressions for the partial derivatives $\partial
%H/\partial q_i$, $\partial H/\partial p_j$ of the system's Hamilton function;
%next, these expressions are programmed and the code is supplied to some
%computational routine implementing a numerical algorithm for ODEs, like Euler,
%Runge-Kutta, etc. Such routines are readily available for virtually all
%programming languages and interactve computational environments.
Standard numerical routines for solving systems of ODEs expect the problem to be
formulated in the form
\begin{equation}\label{eq:s_ode}
  \frac{d\tilde{\psi}}{dt}=\tilde{G}(t, \tilde{\psi})
\end{equation}
where $\tilde\psi$ is an array with the functions to be determined and $\tilde
G$ is a function of time and $\tilde\psi$. In Hamilton's equations for a system
of $F$ degrees of freedom, we have
\begin{equation}\label{eq:hpsi}
  \tilde\psi(t)=[q_1(t), \ldots, q_F(t), p_1(t), \ldots, p_F(t)],\qquad
  H=H(t,\tilde\psi)
\end{equation}
and
\begin{equation}\label{eq:heqsg}
  \tilde G(t, \tilde\psi)=
  \hmatrix
  \tilde\nabla H(t,\tilde{\psi})
\end{equation}
where $\mathbbm{O}$ and $\mathbbm{I}$ are, respectively, the $F\times F$ null
and identity matrices and $\tilde\nabla H$ is a $2F$ componet column array whose
elements are the
partial derivatives of the hamiltonian function in order to the coordinate and 
conjugate momentum variables $q_i$ and $p_j$,
\begin{equation}
  \tilde\nabla H = 
  \begin{pmatrix}
    \partial H/\partial q_1\\
    \vdots\\
    \partial H/\partial p_F
  \end{pmatrix}.
\end{equation}

Usually, expressions for the partial derivatives of the hamiltonian function are
deduced analitically and programmed in suitable form for being repeatedly called
from the computational ODE solver routine. That is, the rhs of eq.
(\ref{eq:s_ode}) is explicitly supplied by the user. Instead, in the method now
proposed, the user only supplies the code for the calculation of hamiltonian
function, and it's up to the numerical ODE solver to compute the hamiltonian's
partial derivatives as needed.

Let us consider a concrete problem and ODE solver to better differenciate the
two alternatives. Take the simple harmonic oscilator, solved using the explicit
forward Euler method~\cite{sir:2010}. The hamiltonian function is
\begin{equation}
  H(q,p) = \frac{1}{2}kq^2+\frac{p^2}{2m},
\end{equation}
where $k$ is the elastic constant of the restoring force and $m$ is the mass of
the oscilator.

%The solution of the system~(\ref{eq:s_ode})
%given~\eqref{eq:hpsi}
%and~\eqref{eq:heqsg} is obtained using the explicit forward Euler
%method~\cite{sir:2010} by iterative application of the formula
%\begin{equation*}
%  \tilde\psi(t+\delta t)=\tilde\psi(t)+\delta t
%  \hmatrix\tilde\nabla H(t,\tilde\psi(t))
%\end{equation*}

@@

\section*{References}
\begin{thebibliography}{99}
  \bibitem{gol:1980} H. Goldstein, \textsl{Classical Mechanics.} Addison-Wesley
  (1980)
  \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in
  Physics.} Springer (2012)
  \bibitem{nr:2007} W.H.~Press, S.A.~Teukolsky, W.T.~Vetterling, B.P.~Flannery,
  \textsl{Numerical Recipes -- The Art of Scientific Computing,} 3rd Ed.
  Cambridge University Press (2007)
\end{thebibliography}
\end{document}

>>>>

\documentclass{iopart}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{bbm}
\expandafter\let\csname equation*\endcsname\relax
\expandafter\let\csname endequation*\endcsname\relax
\usepackage{amsmath}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\undertilde}[1]{\underset{\tilde{}}{#1}}
\newcommand{\hmatrix}{\ensuremath{%
  \begin{pmatrix}
    \mathbbm{O}&\mathbbm{I}\\-\mathbbm{I}&\mathbbm{O}
  \end{pmatrix}
}}
\newcommand{\numap}[1]{\ensuremath{%
    \left\lfloor#1\right\rceil
}}

@@

\submitto{\EJP}
%\maketitle
\section{Introduction}
In the hamiltonian formulation of classical mechanics~\cite{gol:1980}, the time
evolution of physical systems is governed by Hamilton's equations which, for
point particle systems, take the form
\begin{align}\label{eq:heqs}
  \dot q_i&=\pd{H}{p_i}&
  \dot p_i&=-\pd{H}{q_i},
\end{align}
for $i=1,\;\ldots,\;N,$ where $N$ is the number of degrees of freedom of the
system (three times the number of particles for unconstrained, three dimensional
systems), $q_i$, $p_i$ are the canonical coordinates and conjugate momenta
respectively, $H=H(t, q_i, p_i)$ is the hamiltonian function of the system
(for XXX systems, the hamiltonian function is simply the mechanical energy of
the system) and dotted symbols denote their total derivative with respect to
time. For discrete systems, these equations are a set of first order
ordinary differential equations (ODE) in time, whose solutions for given initial
conditions are the trajectories folowed by the system's particles.

@@

%-------------------------------------------------------------------------------
\section{Numerical procedure}
Standard numerical routines for solving systems of ODEs expect the problem to be
written in the form
\begin{equation}\label{eq:numheqs}
  \frac{d\psi_i}{dt}=G_i(t, \psi_1, \psi_2, \ldots),\quad i=1, \ldots, N,
\end{equation}
where $N$ is the number of unknown functions $\psi_i(t)$ (and the number of
equations in well posed problems) and $G_i(t, \psi_1, \psi_2, \ldots)\equiv
G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
the user as callable subprograms. The ODE
solver routine automatically invokes these subprograms in order (as many times
as required by the sspecific method and/or desired accuracy) to compute
estimates of the unknown functions $\psi_i$.

For Hamilton's equations, the $\psi_i$ are the coordinates and momenta, while
the $G_i$ are the partial derivatives of the hamiltonian function, for which, in
the standard approach, the user must derive expressions to be programmed.
Instead, in the method now proposed, only the hamiltonian function is coded and
the ODE solver is modified so that it repeatedly invokes that subprogram in
order to compute the hamiltonian's partial derivatives numerically. Only then
does it proceed to the estimation of the ODE solution.

To make this discussion clearer, let's consider a specific problem, the
classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
a classical oscilator with mass $m$ and restoring force constant $k$ is 
\begin{equation}
  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
\end{equation}
Hamilton's equations~\eqref{eq:heqs} then read
\begin{align}
  \frac{dq}{dt}&=\frac{p}{m}&
  \frac{dp}{dt}&=-kq,
\end{align}
wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
by iterating the formulas
\begin{align}
  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
\end{align}
where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
implements this method, and this the usual procedure. With this paper, we
consider a method where explicit expressions for the hamiltonian's partial
derivatives in the equations of motion are replaced by numerical estimates. In
our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
and~\eqref{eq:efe2} by
\begin{align*}
  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
\end{align*}
where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
In order to make numerical aproximations to the partial derivatives of a
function, one must evaluate it several times; at the very least, two evaluations
are needed. For example, using the central difference method, the derivative is
estimated as
\begin{equation} 
  \frac{\partial\psi}{\partial x} = 
    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
  o(\delta x)^2,
\end{equation}
leading to approximation formulas that, for our example problem, read 
\begin{align}
  \numap{\frac{\partial H}{\partial p}}_i &=
  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
  \numap{\frac{\partial H}{\partial q}}_i &=
  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
\end{align}
where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
Other approximations have higher accuracy, at the cost of extra evaluations of
the function to be derived.

This need for repeated evaluation of the system's hamiltonian reduces the
efficiency of this method. However, programming a procedure for the hamiltonian
function is usually much simpler then doing it for the set of its partial
derivatives.  Also, adjusting the syntax and semantics of the ODE solver and the
hamiltonian procedure is likewise much simpler than when the partial derivatives
are coded.  For small systems, using the hamiltonian is much simpler and faster.

@@

\section*{References}
\begin{thebibliography}{99}
  \bibitem{gol:1980} H. Goldstein, \textsl{Classical Mechanics.} Addison-Wesley
  (1980)
  \bibitem{french:71} A.P.~French, \textsl{Vibration and Waves.} W.W. Norton \&
  Company (1971)
  \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in
  Physics.} Springer (2012)
  \bibitem{nr:2007} W.H.~Press, S.A.~Teukolsky, W.T.~Vetterling, B.P.~Flannery,
  \textsl{Numerical Recipes -- The Art of Scientific Computing,} 3rd Ed.
  Cambridge University Press (2007)
\end{thebibliography}
\end{document}
